A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-02-17 paper/paper.Rmd
updated language on rlm

<!--AE comment: Section 3.2.1 begins “As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.” This could use clarification. In 3.2, you said that part of IDA is “ model formulation without any formal statistical inference.” To me, that sounds like using modeling as a descriptive technique, and it’s something I would consider to be more a part of EDA. But then in 3.2.1, while it is perhaps not using modeling in an inferential way, the model is being used for more than just description. I might rephrase the first sentence as follows: “In order to treat the extreme values in the data, we built a robust linear regression model.” -->
<!--AE comment: Awkward sentence. "which is the model formulation"?--> To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model. We build the model for each ID utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively. This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach. We tried fitting a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP], however perhaps because the data is large and imbalanced, the model was not converging. <!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.

We build the model using the `rlm` function from `MASS` package [@mass]. We set the `mean_hourly_wage` and `year` as the dependent and predictor, respectively. Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm]. <!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. Moreover, it should be noted that not all the outliers are due to an error. Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.

@@

To minimize the risk of mistakenly identifying an outlier as an "erroneous outlier", we simulate some thresholds and study how they affect the data by eye balling the results. Based on this eye balling, we find that 0.12 is the most reasonable value to use as the threshold to minimize the risk of overly smoothing the wage by retaining sensible spikes in the data. In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording. After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called `is_pred`.

@@

```{r compare-plot, fig.cap="Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals. A robust linear model prediction was used to correct mean hourly wages value. We can see that some extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed. Some spikes might remain when wage vaues are plausible.", fig.height=8, fig.width=10, out.width="90%"}
```

>>>>

<!--AE comment: Section 3.2.1 begins “As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.” This could use clarification. In 3.2, you said that part of IDA is “ model formulation without any formal statistical inference.” To me, that sounds like using modeling as a descriptive technique, and it’s something I would consider to be more a part of EDA. But then in 3.2.1, while it is perhaps not using modeling in an inferential way, the model is being used for more than just description. I might rephrase the first sentence as follows: “In order to treat the extreme values in the data, we built a robust linear regression model.” -->
<!--AE comment: Awkward sentence. "which is the model formulation"?--> A robust linear regression model using the `rlm` function from `MASS` package [@mass] is used to treat the extreme values in the data. The robustness weight is used to determine if a value should be replaced with the fitted value from the model. This is constructed for each ID utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively. An alternative approach would be a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP]. However, when we tested this, the fit failed to convergence. <!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.

The `mean_hourly_wage` and `year` are set as the dependent and predictor, respectively. Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm]. <!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. It should be noted that it is not possible to determine if all outliers are errors, and it might be that an individual had abnormally high wages at a particular time.

@@

To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done. Samples of individuals were examined under the different thresholds, with an eye to smoothness of the profiles. Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals. A threshold of 0.12 was chosen, that struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called `is_pred`, so that this change can be monitored in the downstream analyses.

@@

```{r compare-plot, fig.cap="Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for same sample of individuals as shown in Figure 2. A robust linear model prediction was used to identify and correct mean hourly wages value. The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.", fig.height=8, fig.width=10, out.width="90%"}
```

### 2022-02-17 paper/paper.tex
updated language on rlm

To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model. We build the model for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively. This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach. We tried fitting a robust linear mixed model using \texttt{robustlmm} (Koller 2016), however perhaps because the data is large and imbalanced, the model was not converging. The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.

We build the model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002). We set the \texttt{mean\_hourly\_wage} and \texttt{year} as the dependent and predictor, respectively. Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021). The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. Moreover, it should be noted that not all the outliers are due to an error. Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.

To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier'', we simulate some thresholds and study how they affect the data by eye balling the results. Based on this eye balling, we find that 0.12 is the most reasonable value to use as the threshold to minimize the risk of overly smoothing the wage by retaining sensible spikes in the data. In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording. After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called \texttt{is\_pred}.

@@

\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals. A robust linear model prediction was used to correct mean hourly wages value. We can see that some extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed. Some spikes might remain when wage vaues are plausible.}\label{fig:compare-plot}
\end{figure}

>>>>

A robust linear regression model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002) is used to treat the extreme values in the data. The robustness weight is used to determine if a value should be replaced with the fitted value from the model. This is constructed for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively. An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016). However, when we tested this, the fit failed to convergence. The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.

The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively. Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021). The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. It should be noted that it is not possible to determine if all outliers are errors, and it might be that an individual had abnormally high wages at a particular time.

To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done. Samples of individuals were examined under the different thresholds, with an eye to smoothness of the profiles. Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals. A threshold of 0.12 was chosen, that struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called \texttt{is\_pred}, so that this change can be monitored in the downstream analyses.

@@

\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for same sample of individuals as shown in Figure 2. A robust linear model prediction was used to identify and correct mean hourly wages value. The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.}\label{fig:compare-plot}
\end{figure}
